pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-21 article.tex
Example 2 finished writing, still need to create figure for it

We start with an idealised 
birth-death tree of five taxa and a crown age of 10.
By idealized, we mean a phylogeny that is easiest to recover
by the Bayesian inference. A phylogeny that is easy to recover has
the most common branch length distribution imaginable. We
use an idealised tree, as we want to measure the baseline noise; that is,
the lowest error that BEAST2 can possibly make.

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
phylogeny <- ape::read.tree(text = "((A:7.1,(B:2.2,C:2.2):4.9):2.9,(D:4.6,E:4.6):5.4):0;")
\end{lstlisting}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_bd.png}
  \caption{pippo}
\end{figure}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

By default, an alignment is created using a Jukes-Cantor (JC) site model
and a strict clock model. A JC site model assumes that mutations
between all nucloetides are constant at an equal rate. 
A strict clock model assumes that
the mutation rate of all species is equal and constant.
Due to this, we state that the generative model for the alignment is
a JC site model and a strict clock model.

In the second step we state our experiments. In this context, we
define an experiment as a combination of an inference model
and the conditions under which a Bayesian inference is created.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always & no & generative model \\
    \hline
  \end{tabular}
  \caption{Experimental setup to answer the first research question}
\end{table}

If we know how the generative model of the alignment, as we did above,
we can re-use that information. We do still need to specify 
a generative tree prior, which is the tree prior underlying the phylogeny.
In this case, we'll pick a (constant-rate) birth-death model:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
model_select_param <- create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = create_bd_tree_prior()
)
\end{lstlisting}

The third and fourth step have sensible defaults, and we are not
interested in using a twin tree. We can now measure the errors BEAST2
makes when inferring the given phylogeny in this way:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

@@

The resulting figure is shown in figure 2.

\richel{Update figure 2 to use idealized tree}
\giovanni{
  Check "pirouettepackage" repo. 
  I uploaded a function to create the idealized tree that you might like.
}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same
  }
\end{figure}

@@

Here we start with a tree generated from an unknown 
diversification model, that has four taxa and a crown age of 5:

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll re-use the alignment parameters of the previous example 
\richel{TODO: add ref to code}

The second step is to select how an inference model is picked.
We will let the inference model that has the highest evidence to be used
in the Bayesian inference. By default, \verb;pirouette; will use
its full arsenal of 40 inference models, which are all combinations of 4 site 
models, 2 clock models and 5 tree priors.

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

Also here, the third and fourth step have sensible defaults, and we are not
interested in using a twin tree. We can now measure the errors BEAST2
makes when inferring the given phylogeny like this:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  alignment_params = alignment_params,
  model_select_params = model_select_param
)
\end{lstlisting}

@@

The following settings are the same as in the previous section:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters 
\richel{TODO: add ref to code}

>>>>

We start with an idealised Yule (pure-birth) tree 
with five taxa and a crown age of ten.
By idealized, we mean a phylogeny that is easiest to recover
by the Bayesian inference. A phylogeny that is easy to recover has
the most common branch length distribution imaginable. We
use an idealised tree, as we want to measure the baseline noise; that is,
the lowest error that BEAST2 can possibly make.

\begin{lstlisting}[
    language=R, floatplacement=H, frame=single, 
    label = {lst:create_ideal_tree}, 
    caption = {Create an ideal tree}
  ]
phylogeny <- create_ideal_tree(n_taxa = 5, crown_age = 10)
\end{lstlisting}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_bd.png}
  \caption{The ideal tree, as created by listing \ref{lst:create_ideal_tree}}
\end{figure}

@@

\begin{lstlisting}[
    language=R, floatplacement=H, 
    frame=single
    label = {lst:create_alignment}, 
    caption = {Create an alignment}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

By default, an alignment is created using a Jukes-Cantor (JC) site model
and a strict clock model. A JC site model assumes that mutations
between all nucleotides are constant at an equal rate. 
A strict clock model assumes that
the mutation rate of all species is equal and constant.
Due to this, we state that the generative model for the alignment is
a JC site model and a strict clock model.

In the second step we state our experiments. In this context, we
define an experiment as a combination of an inference model
and the conditions under which a Bayesian inference is executed.
Within this example, we specify that our experiment uses the
generative model, will always be run and we ignore the evidence for our model,
for the inference model that is also the generative model (using a
JC site model, strict clock model and Yule tree prior).

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always & no & JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the first research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior
  }
\end{table}

Due to sensible defaults, specifying this
results in:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiments <- list(create_experiment())
\end{lstlisting}

All the \verb;pirouette; arguments are bundled
and checked by \verb;create_pir_params;:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments
)
\end{lstlisting}

Running the experiment:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

@@

The resulting figure is shown in figure \ref{fig:example_1}.

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same
  }
  \label{fig:example_1}
\end{figure}

@@

Here we start with a tree generated from an unknown 
diversification model, that has four taxa and a crown age of five:

\begin{lstlisting}[
  language=R, 
  floatplacement=H, 
  frame=single, 
  label = {lst:unknown_phylogeny},
  caption = "A phylogeny with an unknown diversification model"
]
phylogeny <- ape::read.tree(text = "((A:4, B:4):1, (C:4, D:4):1);")
\end{lstlisting}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll re-use the alignment parameters of the previous example
as shown in listing \ref{lst:create_alignment}.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    candidate & best candidate & yes & JC, strict, Yule \\
    candidate & best candidate & yes & JC, strict, BD \\
    ...       & ...            & ... & ... \\
    candidate & best candidate & yes & GTR, RLN, CCP \\
    candidate & best candidate & yes & GTR, RLN, CEP \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the second research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
\end{table}

In the second step we state our experiments. 
Within this example, all our experiments are candidates,
run only if it is the best candidate, the evidence is measured (ignoring
this will give a helpful error) and we use the full arsenal of 
40 inference models, which are all combinations of 4 site 
models, 2 clock models and 5 tree priors.

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiment <- create_all_experiments()
\end{lstlisting}

Also here, the third (the BEAST2 inference) and fourth (measuring the error)
step have sensible defaults, and we are not
interested in using a twin tree. We can create the complete
\verb;pirouette; parameter set (again) like this:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments
)
\end{lstlisting}

Running the experiment:

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
errors <- pir_run(
  phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

@@

The following settings are the same as in the previous section:
phylogeny (listing \ref{lst:unknown_phylogeny}), alignment parameters, inference model selection,
shared inference parameters and error measuring parameters 
\richel{TODO: add ref to code}
